---
title: "Batch correction in predictive models"
author: "Jennifer"
date: "2023-01-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warnings =FALSE)
```



```{r}
#install.packages("doParallel")
#devtools::install_github("wdl2459/ConQuR")
#devtools::install_github("wdl2459/ConQuR", build_vignettes = TRUE, force=TRUE)


#load packages
library(doParallel) 
library(ConQuR)
library(randomForest)
library(dplyr)
library(caret)
library(glmnet)
library(ROCR)
library(kableExtra)


#set working directory
setwd("~/Desktop/stlp new laptop/Batch effect")

###### Load HIVRC data ######
load(file="HIV_experiment1.Rdata") 
taxa = t(Dat_Genus)
meta = metadat1

# delete gender=2
index = which(meta$gender == 2)
taxa = taxa[-index, ]
meta = meta[-index, ]

meta$Study = factor(meta$Study)
meta$hivstatus = factor(meta$hivstatus)
meta$gender = factor(meta$gender)
meta$race = factor(meta$race)

batch  =  as.numeric(meta$Study) 
batchN = as.numeric(as.factor(batch))-1

N = apply(taxa, 1, sum) 

age = meta$Age
bmi = meta$bmi
gender = meta$gender # factor
race = meta$race # factor
hivstatus = meta$hivstatus # factor, key
logN = log(N)
batchid = factor(batchN)
meta$batchid=batchid



```

## Steps

For each batch in 10 batches:

- Set this batch as the test data and the other 9 batches as the training data

- Use ConQuR to correct the training data –> corrected training tax data

- Train a classifier using the corrected training data, apply the fitted classifier on the corrected training data –> Training AUC

- Use ConQuR to correct the test data [use the corrected training data as the reference batch; do not include outcome variable (which is HIV status in this example) during correction ] –> corrected text tax data

- Apply the classifier on the corrected text data –> Test AUC


# With batch correction
## Random Forest
```{r}
summary(batchid)
record_tab_rf = NULL
record_auc_rf_test = NULL
record_auc_rf_train = NULL
best_ref = NULL
for (i in 0:8) {
  set.seed = 1234
  indx_not_include <- which(meta$batchid == i)
  batchid_train = batchid[-indx_not_include]
  age_train = age[-indx_not_include]
  bmi_train = bmi[-indx_not_include]
  gender_train = gender[-indx_not_include]
  hivstatus_train = hivstatus[-indx_not_include]
  taxa_train = taxa[-indx_not_include, ]
  
  age_test = age[indx_not_include]
  bmi_test = bmi[indx_not_include]
  gender_test = gender[indx_not_include]
  hivstatus_test = hivstatus[indx_not_include]
  taxa_test = taxa[indx_not_include, ]
  
  taxa_train_correct_list = list()
  
  taxa_test_correct_list = list()
  
  record_tab_rf_list = list()
  
  record_auc_rf_test[i+1] = 0
  
  for (ref in 0:9) {
    if(ref != i) {
      taxa_train_correct = ConQuR(tax_tab=taxa_train, batchid=batchid_train,
                                  covariates=data.frame(age_train, bmi_train,
                                                        gender_train, hivstatus_train),
                          batch_ref=as.character(ref)) 
      taxa_train_correct_list[ref + 1] = list(taxa_train_correct)
      # Build a random forest classifier using corrected training data
      temp_train = as.matrix(taxa_train_correct)
      rownames(temp_train) = NULL
      colnames(temp_train) = NULL
      rf_classifier = randomForest(hivstatus_train ~ ., data=temp_train,importance=TRUE)
      # Apply fitted classifier on training data and evaluate the performance
      prediction_for_roc_curve_train <- predict(rf_classifier, temp_train, type="prob")
      pred_train = prediction(prediction_for_roc_curve_train[,2], hivstatus_train)
      auc.perf_train = performance(pred_train, measure="auc")@y.values[[1]]
      
      # Do batch correction on test data, with the corrected training data as the reference batch and do not include the outcome
      newbatchid = append(rep(1, nrow(taxa_test)), rep(0, nrow(taxa_train)))
      newbatchid = factor(newbatchid)
      newtaxa = rbind(taxa_test, taxa_train_correct)
      
      age.new = append(age_test, age_train)
      bmi.new = append(bmi_test, bmi_train)
      gender.new = append(gender_test, gender_train)
      
      taxa_test_correct = ConQuR(tax_tab=newtaxa,batchid=newbatchid,
                                 covariates= data.frame(age.new, bmi.new, gender.new),
                              batch_ref="0")
      
      taxa_test_correct = taxa_test_correct[1:nrow(taxa_test), ]
      
      taxa_test_correct_list[ref + 1] = list(taxa_test_correct)
      
      # Apply fitted classifier on test data and evaluate the performance
      temp_test = as.matrix(taxa_test_correct)
      rownames(temp_test) = NULL
      colnames(temp_test) = NULL
    
      prediction_for_roc_curve_test <- predict(rf_classifier, temp_test, type="prob")
      pred_test = prediction(prediction_for_roc_curve_test[,2], hivstatus_test)
      
      record_tab_rf_list[ref + 1] = list(data.frame(prob= prediction_for_roc_curve_test[, 2], hivstatus_test))
    
      auc.perf_test = performance(pred_test, measure="auc")@y.values[[1]]
      
      if (auc.perf_test > record_auc_rf_test[i+1]) {
        record_auc_rf_test[i + 1] = auc.perf_test
        best_ref[i + 1] = ref
        record_auc_rf_train[i + 1] = auc.perf_train
      }
    }
  }
  
  # Save corrected train tax data and corrected test taxa data (choosing the best reference batch)
  write.table(taxa_train_correct_list[[best_ref[i + 1] + 1]], file=paste("/Users/yuxin/Desktop/MS_BIOSTAT//Research/Batch\ correction/V2/", i, "_train_correct_rf.RData", sep=""))
  
  write.table(taxa_test_correct_list[[best_ref[i+1] + 1]], file=paste("/Users/yuxin/Desktop/MS_BIOSTAT//Research/Batch\ correction/V2/", i, "_test_correct_rf.RData", sep=""))

  record_tab_rf = rbind(record_tab_rf, record_tab_rf_list[[best_ref[i+1] + 1]])
}


# cannot calculate AUC for batch 9 as test set, return error:
## Error in .performance.auc(fpr.stop = 1, predictions = c(`1` = 0.844, `2` = 0.93,  : Not enough distinct predictions to compute area under the ROC curve.
## This batch only has hiv cases

save(record_tab_rf, record_auc_rf_train, record_auc_rf_test, best_ref, file = "hiv_rf.RData")
```

## Lasso logistic regression

```{r}
record_tab_lasso = NULL 
record_auc_lasso_train = NULL
record_auc_lasso_test = NULL
best_ref_lasso = NULL
for (i in 0:8) {
  set.seed = 1234
  indx_not_include <- which(meta$batchid == i)
  batchid_train = batchid[-indx_not_include]
  age_train = age[-indx_not_include]
  bmi_train = bmi[-indx_not_include]
  gender_train = gender[-indx_not_include]
  hivstatus_train = hivstatus[-indx_not_include]
  taxa_train = taxa[-indx_not_include, ]
  
  age_test = age[indx_not_include]
  bmi_test = bmi[indx_not_include]
  gender_test = gender[indx_not_include]
  hivstatus_test = hivstatus[indx_not_include]
  taxa_test = taxa[indx_not_include, ]
  
  #save(taxa_train_correct, file=paste(i, "_tax_train_correct.RData", sep=""))
  
  taxa_train_correct_list = list()
  
  taxa_test_correct_list = list()
  
  record_tab_lasso_list = list()
  
  record_auc_lasso_test[i+1] = 0
  
  for (ref in 0:9) {
    if(ref != i) {
      taxa_train_correct = ConQuR(tax_tab=taxa_train, batchid=batchid_train, covariates=data.frame(age_train, bmi_train, gender_train, hivstatus_train),
                            batch_ref=as.character(ref)) 
      taxa_train_correct_list[ref + 1] = list(taxa_train_correct)
        # Build a random forest classifier using corrected training data
      temp_train = as.matrix(taxa_train_correct)
      rownames(temp_train) = NULL
      colnames(temp_train) = NULL
      
      lasso_classifier = cv.glmnet(x=temp_train, y=hivstatus_train, alpha=1, type.measure='auc', family = "binomial")
      
      # Apply fitted classifier on training data and evaluate the performance
      prediction_for_roc_curve_train <- predict(lasso_classifier, temp_train, type="response", s='lambda.min')
      pred_train = prediction(prediction_for_roc_curve_train, hivstatus_train)
      
      auc.perf_train = performance(pred_train, measure="auc")@y.values[[1]]
      
      # Do batch correction on test data, with the corrected training data as the reference batch and do not include the outcome
      newbatchid = append(rep(1, nrow(taxa_test)), rep(0, nrow(taxa_train)))
      newbatchid = factor(newbatchid)
      newtaxa = rbind(taxa_test, taxa_train_correct)
      
      age.new = append(age_test, age_train)
      bmi.new = append(bmi_test, bmi_train)
      gender.new = append(gender_test, gender_train)
      
      taxa_test_correct = ConQuR(tax_tab=newtaxa, batchid=newbatchid, covariates=data.frame(age.new, bmi.new, gender.new),
                              batch_ref="0")
      
      taxa_test_correct = taxa_test_correct[1:nrow(taxa_test), ]
      
      taxa_test_correct_list[ref + 1] = list(taxa_test_correct)
      
      # Apply fitted classifier on test data and evaluate the performance
      temp_test = as.matrix(taxa_test_correct)
      rownames(temp_test) = NULL
      colnames(temp_test) = NULL
      
      
      prediction_for_roc_curve_test <- predict(lasso_classifier, temp_test, type="response", s='lambda.min')
      pred_test = prediction(prediction_for_roc_curve_test, hivstatus_test)
      
      record_tab_lasso_list[ref + 1] = list(data.frame(prob=prediction_for_roc_curve_test, hivstatus_test))
    
      auc.perf_test = performance(pred_test, measure="auc")@y.values[[1]]
      
      if (auc.perf_test > record_auc_lasso_test[i+1]) {
        record_auc_lasso_test[i + 1] = auc.perf_test
        best_ref_lasso[i + 1] = ref
        record_auc_lasso_train[i + 1] = auc.perf_train
      }
    }
  }
  
  # Save corrected train tax data and corrected test taxa data (choosing the best reference batch)
  write.table(taxa_train_correct_list[[best_ref_lasso[i + 1] + 1]], file=paste("/Users/yuxin/Desktop/MS_BIOSTAT//Research/Batch\ correction/V2/", i, "_train_correct_lasso.RData", sep=""))
  
  write.table(taxa_test_correct_list[[best_ref_lasso[i+1] + 1]], file=paste("/Users/yuxin/Desktop/MS_BIOSTAT//Research/Batch\ correction/V2/", i, "_test_correct_lasso.RData", sep=""))

  record_tab_lasso = rbind(record_tab_lasso, record_tab_lasso_list[[best_ref_lasso[i+1] + 1]])
}


# cannot calculate AUC for batch 9 as test set, return error:
## Error in .performance.auc(fpr.stop = 1, predictions = c(`1` = 0.844, `2` = 0.93,  : Not enough distinct predictions to compute area under the ROC curve.
## This batch only has hiv cases

save(record_tab_lasso, record_auc_lasso_train, record_auc_lasso_test, best_ref_lasso, file = "hiv_lasso.RData")
  
  
```

# Without batch correction
## Random Forest

```{r}
record_tab_rf_ref = NULL
record_auc_rf_ref_train = NULL
record_auc_rf_ref_test = NULL
for (i in 0:8) {
  set.seed = 1234
  indx_not_include <- which(meta$batchid == i)
  batchid_train = batchid[-indx_not_include]
  age_train = age[-indx_not_include]
  bmi_train = bmi[-indx_not_include]
  gender_train = gender[-indx_not_include]
  hivstatus_train = hivstatus[-indx_not_include]
  taxa_train = taxa[-indx_not_include, ]
  
  age_test = age[indx_not_include]
  bmi_test = bmi[indx_not_include]
  gender_test = gender[indx_not_include]
  hivstatus_test = hivstatus[indx_not_include]
  taxa_test = taxa[indx_not_include, ]
  
  # Build a random forest classifier using corrected training data
  temp_train = as.matrix(taxa_train)
  rownames(temp_train) = NULL
  colnames(temp_train) = NULL
  
  rf_classifier = randomForest(hivstatus_train ~ ., data=temp_train, importance=TRUE)
  
  # Apply fitted classifier on training data and evaluate the performance
  prediction_for_roc_curve_train <- predict(rf_classifier, temp_train, type="prob")
  pred_train = prediction(prediction_for_roc_curve_train[,2], hivstatus_train)
  
  auc.perf_train = performance(pred_train, measure="auc")

  record_auc_rf_ref_train[i+1] = auc.perf_train@y.values[[1]]
  
  
  # Apply fitted classifier on test data and evaluate the performance
  temp_test = as.matrix(taxa_test)
  rownames(temp_test) = NULL
  colnames(temp_test) = NULL

  prediction_for_roc_curve_test <- predict(rf_classifier, temp_test, type="prob")
  pred_test = prediction(prediction_for_roc_curve_test[,2], hivstatus_test)
  
  record_tab_rf_ref = rbind(record_tab_rf_ref, data.frame(prob=prediction_for_roc_curve_test[,2], hivstatus_test))

  auc.perf_test = performance(pred_test, measure="auc")

  record_auc_rf_ref_test[i+1] = auc.perf_test@y.values[[1]]
}


# cannot calculate AUC for batch 9 as test set, return error:
## Error in .performance.auc(fpr.stop = 1, predictions = c(`1` = 0.844, `2` = 0.92,  : Not enough distinct predictions to compute area under the ROC curve.
## This batch only has hiv cases

save(record_tab_rf_ref, record_auc_rf_ref_train, record_auc_rf_ref_test, best_ref, file = "hiv_rf_ref.RData")


```

## Lasso logistic regression
```{r}

record_tab_lasso_ref = NULL
record_auc_lasso_ref_train = NULL
record_auc_lasso_ref_test = NULL
for (i in 0:9) {
  set.seed = 1234
  indx_not_include <- which(meta$batchid == i)
  batchid_train = batchid[-indx_not_include]
  age_train = age[-indx_not_include]
  bmi_train = bmi[-indx_not_include]
  gender_train = gender[-indx_not_include]
  hivstatus_train = hivstatus[-indx_not_include]
  taxa_train = taxa[-indx_not_include, ]
  
  age_test = age[indx_not_include]
  bmi_test = bmi[indx_not_include]
  gender_test = gender[indx_not_include]
  hivstatus_test = hivstatus[indx_not_include]
  taxa_test = taxa[indx_not_include, ]
  
  # Build a random forest classifier using corrected training data
  temp_train = as.matrix(taxa_train)
  rownames(temp_train) = NULL
  colnames(temp_train) = NULL
  
  lasso_classifier = cv.glmnet(x=temp_train, y=hivstatus_train, alpha=1, type.measure='auc', family = "binomial")
  
  # Apply fitted classifier on training data and evaluate the performance
  prediction_for_roc_curve_train <- predict(lasso_classifier, temp_train, type="response", s='lambda.min')
  pred_train = prediction(prediction_for_roc_curve_train, hivstatus_train)
  
  auc.perf_train = performance(pred_train, measure="auc")

  record_auc_lasso_ref_train[i+1] = auc.perf_train@y.values[[1]]
  
  # Apply fitted classifier on test data and evaluate the performance
  temp_test = as.matrix(taxa_test)
  rownames(temp_test) = NULL
  colnames(temp_test) = NULL

  prediction_for_roc_curve_test <- predict(lasso_classifier, temp_test, type="response", s='lambda.min')
  pred_test = prediction(prediction_for_roc_curve_test, hivstatus_test)
  
  record_tab_lasso_ref = rbind(record_tab_lasso_ref, data.frame(prob=prediction_for_roc_curve_test, hivstatus_test))

  auc.perf_test = performance(pred_test, measure="auc")

  record_auc_lasso_ref_test[i+1] = auc.perf_test@y.values[[1]]
}


# cannot calculate AUC for batch 9 as test set, return error:
## Error in .performance.auc(fpr.stop = 1, predictions = c(`1` = 0.844, `2` = 0.92,  : Not enough distinct predictions to compute area under the ROC curve.
## This batch only has hiv cases

save(record_tab_lasso_ref, record_auc_lasso_ref_train, record_auc_lasso_ref_test, file = "hiv_lasso_ref.RData")

```


# Results

## Random Forest

```{r}


load(file="hiv_rf.RData")
load(file="hiv_rf_ref.RData")
# record_auc_rf_train
# record_auc_rf_test
record_auc_rf_train[10] = NA
record_auc_rf_test[10] = NA
record_auc_rf_ref_train[10] = NA
record_auc_rf_ref_test[10] = NA
best_ref[10] = NA

rf_dat <- data.frame(cbind(record_auc_rf_train, record_auc_rf_test, best_ref, record_auc_rf_ref_train, record_auc_rf_ref_test))
colnames(rf_dat)<-c("Training AUC", "Test AUC", "Chosen Reference Batch", "Training AUC", "Test AUC")
rownames(rf_dat)<-c("Batch 0", "Batch 1", "Batch 2", "Batch 3", "Batch 4", "Batch 5", "Batch 6", "Batch 7", "Batch 8", "Batch 9")
kbl(rf_dat) %>%
  kable_classic() %>%
  add_header_above(c(" " = 1, "Batch corrected" = 3, "Reference" = 2))


```

## Lasso logistic regression

```{r}

load(file="hiv_lasso.RData")
load(file="hiv_lasso_ref.RData")

record_auc_lasso_train[10] = NA
record_auc_lasso_test[10] = NA
record_auc_lasso_ref_train[10] = NA
record_auc_lasso_ref_test[10] = NA
best_ref_lasso[10] = NA

lasso_dat <- data.frame(cbind(record_auc_lasso_train, record_auc_lasso_test, best_ref_lasso, record_auc_lasso_ref_train, record_auc_lasso_ref_test))
colnames(lasso_dat)<-c("Training AUC", "Test AUC", "Chosen Reference Batch", "Training AUC", "Test AUC")
rownames(lasso_dat)<-c("Batch 0", "Batch 1", "Batch 2", "Batch 3", "Batch 4", "Batch 5", "Batch 6", "Batch 7", "Batch 8", "Batch 9")
kbl(lasso_dat) %>%
  kable_classic() %>%
  add_header_above(c(" " = 1, "Batch corrected" = 3, "Reference" = 2))
```







